---
title: "STA 9750 Mini Project #01"
author: "Izabelle Diaz"
format: 
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Introduction

In 2023, the U.S. Supreme Court ruled in *Students for Fair Admissions v. Harvard* (SFFA) that race-conscious admissions policies violated the Equal Protection Clause of the Fourteenth Amendment. This decision significantly changed how universities approach admissions and raised questions about how the ruling might affect student diversity across institutions.

This project analyzes admissions and enrollment data from the Integrated Postsecondary Education Data System (IPEDS), maintained by the U.S. Department of Education. The dataset provides detailed information on college enrollment, including demographic characteristics of students across institutions.

Using data analysis tools in R, this report examines enrollment patterns and demographic trends in higher education institutions before and after the SFFA decision. The goal is to explore whether the removal of race-conscious admissions policies has had any observable impact on the composition of student populations at universities across the United States.

---

## Setup

```{r}
# Load required libraries
library(tidyverse)
library(readr)
library(stringr)
library(ggplot2)
library(gt)

# Load course helper functions
source("https://michael-weylandt.com/STA9750/load_helpers.R")
```

## Data Acquisition

The following code downloads enrollment and institutional data from the Integrated Postsecondary Education Data System (IPEDS). To make the process efficient, the data are saved locally in the `data/mp01` folder so the files do not need to be downloaded again if the analysis is rerun.

```{r}
# Acquire IPEDS Data for MP#01

acquire_ipeds_data <- function(start_year=2010, end_year=2024){
    library(tidyverse)
    library(glue)
    
    data_dir <- file.path("data", "mp01")
    
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings=FALSE, recursive=TRUE)
    }
    
    YEARS <- seq(start_year, end_year)
    
    EFA_ALL <- map(YEARS, function(yy){
        if(yy <= 2022){
            ef_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/EF{yy}A.zip")
        } else {
            ef_url <- glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=EF{yy}A&HasRV=0&type=csv")
        }
        
        ef_file <- file.path(data_dir, glue("ef{yy}a.csv.zip"))
        
        if(!file.exists(ef_file)){
            message(glue("Downloading Enrollment Data for {yy}"))
            download.file(ef_url, destfile = ef_file, quiet=TRUE, mode="wb")    
        }
        
        read_csv(ef_file, show_col_types=FALSE) |>
            mutate(
                institution_id = UNITID,
                year = yy
            ) |>
            select(institution_id, year, everything())
        
    }) |> bind_rows()
    
    DESC_ALL <- map(YEARS, function(yy){
        if(yy <= 2022){
            hd_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/HD{yy}.zip")
        } else {
            hd_url <- glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=HD{yy}&HasRV=0&type=csv")
        }
        
        hd_file <- file.path(data_dir, glue("hd{yy}.csv.zip"))
        
        if(!file.exists(hd_file)){
            message(glue("Downloading Institutional Data for {yy}"))
            download.file(hd_url, destfile = hd_file, quiet=TRUE, mode="wb")    
        }
        
        read_csv(hd_file, show_col_types=FALSE) |>
            mutate(year = yy) |>
            select(
                institution_id = UNITID,
                institution_name = INSTNM,
                state = STABBR,
                year
            )
        
    }) |> bind_rows()
    
    inner_join(EFA_ALL, DESC_ALL, by = c("institution_id","year"))
}

IPEDS <- acquire_ipeds_data()
```

## Task 2: Identify CUNY Schools

```{r}
IPEDS <- IPEDS |>
  mutate(
    is_cuny = str_detect(institution_name, "CUNY")
  )
```

## Task 3: Identify UC and Cal State Schools

```{r}
IPEDS <- IPEDS |>
  mutate(
    is_calpublic =
      str_detect(institution_name, "University of California") |
      str_detect(institution_name, "California State University")
  )

IPEDS |>
  filter(is_calpublic == TRUE) |>
  distinct(institution_name) |>
  arrange(institution_name)
```

## Task 4: Initial EDA Pass

Before beginning deeper analysis, it is helpful to inspect the dataset to understand its structure and verify that the variables were imported correctly.

### Inspecting the structure of the dataset

```{r}
glimpse(IPEDS)
```

### Summary statistics

```{r}
summary(IPEDS)
```

### Observations

The `glimpse()` output shows the structure of the IPEDS dataset, including the number of rows and columns as well as the data type of each variable. The dataset contains information about institutions, enrollment counts, demographic categories, reporting year, and institutional characteristics such as whether a school belongs to the CUNY system or California public systems.

The `summary()` output provides basic statistics for the variables in the dataset. Numeric variables such as enrollment counts show ranges of values across institutions and years, while categorical variables display the frequency of different categories. These checks confirm that the dataset has been loaded correctly and that the variables appear to be structured appropriately for further analysis.

## Task 5: Exploratory Questions - Inline Values

```{r}
library(scales)

IPEDS <- IPEDS |>
  mutate(
    total_students =
      EFAIANM + EFAIANW +
      EFASIAM + EFASIAW +
      EFBKAAM + EFBKAAW +
      EFHISPM + EFHISPW +
      EFNHPIM + EFNHPIW +
      EFWHITM + EFWHITW +
      EF2MORM + EF2MORW +
      EFUNKNM + EFUNKNW +
      EFNRALM + EFNRALW,

    female_students =
      EFAIANW + EFASIAW + EFBKAAW +
      EFHISPW + EFNHPIW + EFWHITW +
      EF2MORW + EFUNKNW + EFNRALW,

    nhpi_students = EFNHPIM + EFNHPIW
  )
```

There are `r comma(n_distinct(IPEDS$institution_id))` distinct institutions represented in this dataset.

In 2024, Baruch College had `r comma(IPEDS |>
  filter(str_detect(institution_name, "Baruch"),
         year == 2024) |>
  summarise(sum(total_students, na.rm = TRUE)) |>
  pull())` students enrolled.

The institution with the highest number of female students in 2019 was `r IPEDS |>
  filter(year == 2019) |>
  group_by(institution_name) |>
  summarise(total_female = sum(female_students, na.rm = TRUE)) |>
  arrange(desc(total_female)) |>
  slice(1) |>
  pull(institution_name)` with `r comma(IPEDS |>
  filter(year == 2019) |>
  group_by(institution_name) |>
  summarise(total_female = sum(female_students, na.rm = TRUE)) |>
  arrange(desc(total_female)) |>
  slice(1) |>
  pull(total_female))` female students.

Among institutions with more than 1,000 students, the institution with the highest proportion of Native Hawaiian or Pacific Islander students in 2024 was `r IPEDS |>
  filter(year == 2024) |>
  group_by(institution_name) |>
  summarise(total = sum(total_students, na.rm = TRUE),
            nhpi = sum(nhpi_students, na.rm = TRUE)) |>
  filter(total > 1000) |>
  mutate(frac = nhpi / total) |>
  arrange(desc(frac)) |>
  slice(1) |>
  pull(institution_name)` with a proportion of `r percent(IPEDS |>
  filter(year == 2024) |>
  group_by(institution_name) |>
  summarise(total = sum(total_students, na.rm = TRUE),
            nhpi = sum(nhpi_students, na.rm = TRUE)) |>
  filter(total > 1000) |>
  mutate(frac = nhpi / total) |>
  arrange(desc(frac)) |>
  slice(1) |>
  pull(frac))`.
  
  