---
title: "STA 9750 Mini Project #01"
author: "Izabelle Diaz"
format: 
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Introduction

In 2023, the U.S. Supreme Court ruled in *Students for Fair Admissions v. Harvard* (SFFA) that race-conscious admissions policies violated the Equal Protection Clause of the Fourteenth Amendment. This decision significantly changed how universities approach admissions and raised questions about how the ruling might affect student diversity across institutions.

This project analyzes admissions and enrollment data from the Integrated Postsecondary Education Data System (IPEDS), maintained by the U.S. Department of Education. The dataset provides detailed information on college enrollment, including demographic characteristics of students across institutions.

Using data analysis tools in R, this report examines enrollment patterns and demographic trends in higher education institutions before and after the SFFA decision. The goal is to explore whether the removal of race-conscious admissions policies has had any observable impact on the composition of student populations at universities across the United States.

---

## Setup

```{r}
# Load required libraries
library(tidyverse)
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(gt)

# Load course helper functions
source("https://michael-weylandt.com/STA9750/load_helpers.R")
```

## Data Acquisition

The following code downloads enrollment and institutional data from the Integrated Postsecondary Education Data System (IPEDS). To make the process efficient, the data are saved locally in the `data/mp01` folder so the files do not need to be downloaded again if the analysis is rerun.

```{r}
# Acquire IPEDS Data for MP#01

acquire_ipeds_data <- function(start_year=2010, end_year=2024){
    library(tidyverse)
    library(glue)
    
    data_dir <- file.path("data", "mp01")
    
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings=FALSE, recursive=TRUE)
    }
    
    YEARS <- seq(start_year, end_year)
    
    EFA_ALL <- map(YEARS, function(yy){
        if(yy <= 2022){
            ef_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/EF{yy}A.zip")
        } else {
            ef_url <- glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=EF{yy}A&HasRV=0&type=csv")
        }
        
        ef_file <- file.path(data_dir, glue("ef{yy}a.csv.zip"))
        
        if(!file.exists(ef_file)){
            message(glue("Downloading Enrollment Data for {yy}"))
            download.file(ef_url, destfile = ef_file, quiet=TRUE, mode="wb")    
        }
        
        read_csv(ef_file, show_col_types=FALSE) |>
            mutate(
                institution_id = UNITID,
                year = yy
            ) |>
            select(institution_id, year, everything())
        
    }) |> bind_rows()
    
    DESC_ALL <- map(YEARS, function(yy){
        if(yy <= 2022){
            hd_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/HD{yy}.zip")
        } else {
            hd_url <- glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=HD{yy}&HasRV=0&type=csv")
        }
        
        hd_file <- file.path(data_dir, glue("hd{yy}.csv.zip"))
        
        if(!file.exists(hd_file)){
            message(glue("Downloading Institutional Data for {yy}"))
            download.file(hd_url, destfile = hd_file, quiet=TRUE, mode="wb")    
        }
        
        read_csv(hd_file, show_col_types=FALSE) |>
            mutate(year = yy) |>
            select(
                institution_id = UNITID,
                institution_name = INSTNM,
                state = STABBR,
                year
            )
        
    }) |> bind_rows()
    
    inner_join(EFA_ALL, DESC_ALL, by = c("institution_id","year"))
}

IPEDS <- acquire_ipeds_data()
```

## Task 2: Identify CUNY Schools

```{r}
IPEDS <- IPEDS |>
  mutate(
    is_cuny = str_detect(institution_name, "CUNY")
  )
```

## Task 3: Identify UC and Cal State Schools

```{r}
IPEDS <- IPEDS |>
  mutate(
    is_calpublic =
      str_detect(institution_name, "University of California") |
      str_detect(institution_name, "California State University")
  )

IPEDS |>
  filter(is_calpublic == TRUE) |>
  distinct(institution_name) |>
  arrange(institution_name)
```
