---
title: "STA 9750 Mini Project 02"
author: "Izabelle Diaz"
format:
  html:
    theme: sandstone
---

## Task 1: Data Acquisition

```{r}
library(tidyverse)
library(glue)
library(readxl)
library(httr2)

load_atus_data <- function(file_base = c("resp", "rost", "act")) {
    
    if (!dir.exists(file.path("data", "mp02"))) {
        dir.create(file.path("data", "mp02"), showWarnings = FALSE, recursive = TRUE)
    }

    file_base <- match.arg(file_base)

    file_name_out <- file.path(
        "data", "mp02",
        glue("atus{file_base}_0324.dat")
    )

    if (!file.exists(file_name_out)) {
        url_end <- glue("atus{file_base}-0324.zip")
        temp_zip <- tempfile(fileext = ".zip")
        temp_dir <- tools::file_path_sans_ext(temp_zip)

        request("https://www.bls.gov") |>
            req_url_path("tus", "datafiles", url_end) |>
            req_headers(`User-Agent` = "Mozilla/5.0") |>
            req_perform(path = temp_zip)

        unzip(temp_zip, exdir = temp_dir)

        temp_file <- file.path(temp_dir, basename(file_name_out))
        file.copy(temp_file, file_name_out, overwrite = TRUE)
    }

    file <- read_csv(file_name_out, show_col_types = FALSE)

    result <- switch(
        file_base,

        resp = file |>
            mutate(
                children_under_18 = TRCHILDNUM,
                employment_status = case_match(
                    TELFS,
                    1 ~ "Employed - at work",
                    2 ~ "Employed - absent",
                    3 ~ "Unemployed - on layoff",
                    4 ~ "Unemployed - looking",
                    5 ~ "Not in labor force",
                    .default = "Other"
                )
            ) |>
            rename(
                survey_weight = TUFNWGTP,
                time_alone = TRTALONE,
                survey_year = TUYEAR
            ),

        rost = file,

        act = file |>
            mutate(
                time_spent_min = TUACTDUR24,
                level3_code = TRCODEP
            )
    )

    result |>
        rename(participant_id = TUCASEID)
}

atus_resp <- load_atus_data("resp")
atus_act  <- load_atus_data("act")

atus_activity_labels <- read_csv(
  "https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv",
  show_col_types = FALSE
)
```

The ATUS data was loaded using the instructor-provided functions for the respondent and activity files, along with the activity code lookup table. This created three datasets: `atus_resp`, `atus_act`, and `atus_activity_labels`.

The respondent dataset contains demographic and survey-level information for each participant, including variables such as survey year and survey weight. The activity dataset contains detailed time-use records, where each row represents one activity completed by a participant along with its duration and activity code. The activity labels dataset provides readable names for activity codes.

Based on the structure of these datasets, the main key used to connect them is `participant_id`. The activity dataset can also be joined with the activity labels dataset using activity codes to translate numeric values into meaningful activity names.

---

## Task 2: Practice Joining Data Files

```{r}
sleep_total_hours <- atus_act |>
    inner_join(atus_activity_labels, join_by(level3_code == code_level3)) |>
    filter(str_detect(tolower(task_level3), "sleep")) |>
    summarise(total_hours = sum(time_spent_min, na.rm = TRUE) / 60)

sleep_total_hours

female_no_time_alone_2003 <- atus_resp |>
    filter(survey_year == 2003, time_alone == 0) |>
    summarise(total_participants = n())

female_no_time_alone_2003
```

For the first question, the activity table provides the time values, while the activity label table provides the meaning of the activity codes, so a join is required. For the second question, no join is needed because the respondent table already contains survey year and time alone.

---

## Task 3: Adding Additional Demographic Variables

To complete Task 3, I modified the data loading function to include additional demographic variables such as the number of children under 18 and employment status. The number of children was captured using `TRCHILDNUM`, while employment status was derived from `TELFS`, which classifies respondents based on their labor force participation.

These additional variables help provide more context for analyzing how individuals spend their time across different demographic groups.

---

## Task 4: Exploratory Questions - Inline Values

```{r}
total_respondents <- n_distinct(atus_resp$participant_id)

sleep_hours <- atus_act |>
  inner_join(atus_activity_labels, join_by(level3_code == code_level3)) |>
  filter(str_detect(tolower(task_level3), "sleep")) |>
  summarise(hours = mean(time_spent_min, na.rm = TRUE) / 60) |>
  pull(hours)
```

Since 2003, `r total_respondents` different respondents have completed an ATUS survey.

On average, Americans sleep about `r round(sleep_hours, 2)` hours per day.

---

## Task 5: Exploratory Questions - Table Answers

```{r}
library(gt)

lawn_care <- atus_act |>
  inner_join(atus_activity_labels, join_by(level3_code == code_level3)) |>
  inner_join(atus_resp, join_by(participant_id)) |>
  filter(str_detect(tolower(task_level3), "lawn")) |>
  mutate(group = case_when(
    employment_status == "Not in labor force" ~ "Not working",
    employment_status == "Employed - at work" ~ "Working",
    TRUE ~ "Other"
  )) |>
  filter(group %in% c("Not working", "Working")) |>
  group_by(group) |>
  summarise(avg_minutes = mean(time_spent_min, na.rm = TRUE)) |>
  mutate(avg_hours = avg_minutes / 60)

lawn_care |> gt()
```

---

## Task 6: Exploratory Questions - Graph Answers

```{r}
simple_plot <- atus_act |>
  inner_join(atus_activity_labels, by = c("level3_code" = "code_level3")) |>
  count(task_level1, sort = TRUE) |>
  slice_head(n = 10)

ggplot(simple_plot, aes(x = reorder(task_level1, n), y = n)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Most Common Activity Categories",
    x = "Activity Type",
    y = "Count"
  )
```

As shown in the plot, the most common activities reported by Americans are concentrated in a few broad categories, especially personal care, socializing and leisure, and household activities. This suggests that a large share of daily time use is devoted to routine and essential parts of everyday life rather than less frequent or specialized tasks.

---

## Task 7: Time Use Categories for Target Demographics

### Stay-at-Home Dads

The first target demographic is stay-at-home fathers. This group is defined as adult men who are not in the labor force and have at least one child under the age of 18 in the household. These individuals are likely responsible for childcare and household responsibilities.

Key activities for this group include childcare, household work, and personal time such as sleep. Based on the data, childcare and household activities take up a significant portion of their day.

Typical time use levels:  
- Childcare: Low = 1 hour, Medium = 3 hours, High = 6 hours  
- Household activities: Low = 1 hour, Medium = 2.5 hours, High = 5 hours  
- Sleep: Low = 5 hours, Medium = 7 hours, High = 9 hours  

---

### College Undergraduates

The second demographic is college undergraduate students. This group is defined as individuals who are currently enrolled in college-level education programs and are typically between ages 18 and 24.

Key activities include studying, attending class, sleeping, and leisure activities. While students spend time on education, there is also a significant amount of time allocated to personal and recreational activities.

Typical time use levels:  
- Studying: Low = 1 hour, Medium = 3 hours, High = 6 hours  
- Sleep: Low = 5 hours, Medium = 7 hours, High = 9 hours  
- Leisure: Low = 2 hours, Medium = 4 hours, High = 7 hours  

---

### Young Working Professionals

The final demographic is young working professionals. This group is defined as individuals between the ages of 22 and 30 who are employed full-time and not currently enrolled in school.

Key activities include work, exercise, sleep, and leisure. This group often balances career development with personal well-being.

Typical time use levels:  
- Work: Low = 6 hours, Medium = 8 hours, High = 10+ hours  
- Exercise: Low = 0.5 hours, Medium = 1 hour, High = 2 hours  
- Sleep: Low = 5 hours, Medium = 7 hours, High = 8+ hours  

---
<!--
::: {.callout-note title="AI Usage Statement" collapse="true"}

I used ChatGPT to help debug R errors, understand Quarto formatting, and structure my code for data loading, joining, and summarizing. I reviewed all outputs and made sure I understood the logic behind each step before including it in my submission.

:::
-->

<!--
::: {.callout-note title="AI Usage Statement" collapse="true"}

I used ChatGPT to help debug R errors, understand Quarto formatting, and structure my code for data loading, joining, and summarizing. I reviewed all outputs and made sure I understood the logic behind each step before including it in my submission.

:::
-->

<div style="border-left:6px solid #3b82f6; background-color:#eff6ff; padding:12px 16px; border-radius:6px;">

<details>
<summary><strong>AI Usage Statement (click to expand)</strong></summary>

<br>

I used ChatGPT to help debug R errors, understand Quarto formatting, and structure my code for data loading, joining, and summarizing. I reviewed all outputs and made sure I understood the logic behind each step before including it in my submission.

</details>

</div>