---
title: "STA 9750 Mini Project 03"
author: "Izabelle Diaz"
format:
  html:
    embed-resources: true
---

## Task 1: Data Acquisition - State Population

::: {.callout-note collapse="true" title="Show Code"}
```{r}
library(tidyverse)
library(tidycensus)
library(sf)
library(readxl)
library(jsonlite)
library(gt)
library(scales)

dir.create("data/mp03", recursive = TRUE, showWarnings = FALSE)

state_pop_2015_2024 <- map_dfr(c(2015:2019, 2021:2024), function(y) {
  get_acs(
    geography = "state",
    variables = c(total_population = "B01003_001"),
    year = y,
    survey = "acs1"
  ) %>%
    select(NAME, total_population = estimate) %>%
    mutate(year = y)
})

state_population_tidy <- state_pop_2015_2024 %>%
  transmute(
    state = NAME,
    year,
    total_population
  )
```
:::

To obtain baseline population data, I used the `get_acs()` function from the `tidycensus` package to retrieve state-level total population estimates from 2015 to 2024, excluding 2020 because the standard ACS 1-year estimates were not released for that year. I then cleaned the results into a tidy dataset with the variables `state`, `year`, and `total_population`.

### Preview of Data

```{r}
head(state_population_tidy)
```

## Task 2: Data Acquisition - State-to-State Flows (2024)

::: {.callout-note collapse="true" title="Show Code"}
```{r}
state_to_state_migration_2024 <- function() {

  file_path <- "data/mp03/State_to_State_Migration_Table_2024_T13.xlsx"

  if (!file.exists(file_path)) {
    download.file(
      "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx",
      file_path,
      mode = "wb"
    )
  }

  state_lookup <- c(
    setNames(state.abb, state.name),
    "District of Columbia" = "DC"
  )

  movers_2024 <- read_excel(
    file_path,
    sheet = "Table",
    skip = 8,
    col_names = c("state_current", "state_1y", "population", "moe")
  ) %>%
    mutate(
      state_current = str_trim(as.character(state_current)),
      state_1y = str_trim(as.character(state_1y)),
      population = str_trim(as.character(population)),
      population = replace_na(population, "0"),
      population = replace(population, population == "N", "0")
    ) %>%
    filter(
      !is.na(state_current),
      !is.na(state_1y),
      population != "X",
      state_current != "District of Columbia",
      state_1y != "District of Columbia"
    ) %>%
    mutate(
      population = suppressWarnings(as.numeric(population)),
      year_current = 2024,
      year_1y = 2023,
      state_current = unname(state_lookup[state_current]),
      state_1y = unname(state_lookup[state_1y])
    ) %>%
    filter(
      !is.na(population),
      !is.na(state_current),
      !is.na(state_1y)
    ) %>%
    select(state_current, state_1y, population, year_current, year_1y)

  stationary_2024 <- read_excel(
    file_path,
    sheet = "Supplemental - Current Res",
    skip = 10,
    col_names = FALSE
  ) %>%
    select(
      state_current = 1,
      same_residence = 4,
      same_state = 6
    ) %>%
    filter(!is.na(state_current), state_current != "District of Columbia") %>%
    mutate(
      state_current = str_trim(as.character(state_current)),
      state_1y = state_current,
      same_residence = suppressWarnings(as.numeric(same_residence)),
      same_state = suppressWarnings(as.numeric(same_state)),
      population = replace_na(same_residence, 0) + replace_na(same_state, 0),
      year_current = 2024,
      year_1y = 2023,
      state_current = unname(state_lookup[state_current]),
      state_1y = unname(state_lookup[state_1y])
    ) %>%
    filter(
      !is.na(state_current),
      !is.na(state_1y)
    ) %>%
    select(state_current, state_1y, population, year_current, year_1y)

  bind_rows(movers_2024, stationary_2024)
}
```
:::

To import the 2024 state-to-state migration data, I constructed a function that downloads the Census Excel file if it is not already stored locally and then reshapes the data into a tidy format. I used the main migration sheet for interstate flows and the supplemental sheet to capture residents who remained in the same state.

### Preview of Data

```{r}
migration_2024 <- state_to_state_migration_2024()
head(migration_2024)
dim(migration_2024)
```

## Task 3: Data Acquisition - State-to-State Flows (2023)

::: {.callout-note collapse="true" title="Show Code"}
```{r}
state_to_state_migration_2023 <- function() {

  file_path <- "data/mp03/State_to_State_Migration_Table_2023_T13.xlsx"

  if (!file.exists(file_path)) {
    download.file(
      "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2023/state-to-state-migration/State_to_State_Migration_Table_2023_T13.xlsx",
      file_path,
      mode = "wb"
    )
  }

  states_with_dc <- c(
    state.name[1:8],
    "District of Columbia",
    state.name[9:50]
  )

  state_lookup <- c(
    setNames(state.abb, state.name),
    "District of Columbia" = "DC"
  )

  stayed_data <- read_excel(
    file_path,
    range = readxl::cell_cols("A:F"),
    col_names = FALSE
  ) %>%
    slice(-(1:8)) %>%
    select(
      state_to = 1,
      same_house = 4,
      same_state = 6
    ) %>%
    filter(!is.na(state_to)) %>%
    mutate(
      state_to = str_trim(as.character(state_to)),
      same_house = suppressWarnings(as.numeric(same_house)),
      same_state = suppressWarnings(as.numeric(same_state))
    ) %>%
    filter(state_to %in% state.name) %>%
    mutate(
      state_from = state_to,
      population = replace_na(same_house, 0) + replace_na(same_state, 0),
      year_current = 2023,
      year_1y = 2022,
      state_to = unname(state_lookup[state_to]),
      state_from = unname(state_lookup[state_from])
    ) %>%
    filter(state_to != "DC", state_from != "DC") %>%
    select(state_to, state_from, population, year_current, year_1y)

  flow_data <- read_excel(
    file_path,
    range = readxl::cell_cols("J:DG"),
    col_names = FALSE
  ) %>%
    slice(1:51) %>%
    select(1:51) %>%
    setNames(states_with_dc) %>%
    mutate(state_to = states_with_dc) %>%
    pivot_longer(
      cols = -state_to,
      names_to = "state_from",
      values_to = "population"
    ) %>%
    mutate(
      state_to = str_trim(as.character(state_to)),
      state_from = str_trim(as.character(state_from)),
      population = str_trim(as.character(population))
    ) %>%
    filter(
      !is.na(population),
      population != "N",
      population != "X",
      state_to != "District of Columbia",
      state_from != "District of Columbia",
      state_to != state_from
    ) %>%
    mutate(
      population = suppressWarnings(as.numeric(population)),
      year_current = 2023,
      year_1y = 2022,
      state_to = unname(state_lookup[state_to]),
      state_from = unname(state_lookup[state_from])
    ) %>%
    filter(
      !is.na(population),
      !is.na(state_to),
      !is.na(state_from)
    ) %>%
    select(state_to, state_from, population, year_current, year_1y)

  bind_rows(stayed_data, flow_data) %>%
    rename(
      state_current = state_to,
      state_1y = state_from
    )
}
```
:::

To construct the 2023 state-to-state migration dataset, I imported the within-state counts and the interstate flow section separately. I then cleaned the data, converted it into a long format, removed District of Columbia observations, and combined the pieces into one tidy table.

### Preview of Data

```{r}
migration_2023 <- state_to_state_migration_2023()
head(migration_2023)
dim(migration_2023)
```

## Task 4: Data Acquisition - Combined State-to-State Flows

::: {.callout-note collapse="true" title="Show Code"}
```{r}
state_to_state_migration <- function() {
  flow_2024 <- state_to_state_migration_2024()
  flow_2023 <- state_to_state_migration_2023()

  bind_rows(flow_2024, flow_2023)
}
```
:::

I then combined the 2023 and 2024 state migration datasets into one larger table for later analysis.

### Preview of Data

```{r}
state_flows <- state_to_state_migration()
head(state_flows)
dim(state_flows)
```

## Task 5: Data Acquisition - Metro-to-Metro Migration Flows

::: {.callout-note collapse="true" title="Show Code"}
```{r}
metro_to_metro_migration <- function() {
  library(tidyverse)
  library(jsonlite)
  
  dir.create("data/mp03", recursive = TRUE, showWarnings = FALSE)
  
  file_path <- "data/mp03/metro_to_metro_migration_2023.json"
  
  api_url <- paste0(
    "https://api.census.gov/data/2020/acs/flows?",
    "get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME,GEOID1,GEOID2&",
    "for=metropolitan%20statistical%20area/micropolitan%20statistical%20area:*"
  )
  
  if (!file.exists(file_path)) {
    download.file(
      url = api_url,
      destfile = file_path,
      mode = "wb"
    )
  }
  
  move_row_to_colnames <- function(X, row = 1) {
    X <- as.data.frame(X, stringsAsFactors = FALSE)
    Xrow <- X[row, , drop = FALSE]
    X <- X[-row, , drop = FALSE]
    colnames(X) <- unlist(Xrow[1, ])
    tibble::as_tibble(X) |>
      readr::type_convert()
  }
  
  pull_state_from_metro <- function(metro_name) {
    stringr::str_extract(metro_name, "(?<=, )[A-Z]{2}(?=-[A-Z]{2}|- Metro Area)")
  }
  
  metro_data <- read_json(file_path, simplifyVector = TRUE) |>
    move_row_to_colnames() |>
    transmute(
      metro_current = FULL1_NAME,
      metro_1y = FULL2_NAME,
      moved_in = MOVEDIN,
      moved_out = MOVEDOUT,
      state_current = pull_state_from_metro(metro_current),
      state_1y = pull_state_from_metro(metro_1y)
    )
  
  metro_data
}
```
:::

To import metro-to-metro migration flows, I created a function called `metro_to_metro_migration()` that retrieves data from the Census API and saves it locally as a JSON file. This ensures the data is only downloaded once and can be reused efficiently.

Because the raw API output does not include proper column names, I used a helper function (`move_row_to_colnames`) to convert the first row of the dataset into column names and then cleaned the data structure. I also extracted state abbreviations from metro area names using a regular expression.

The final dataset includes metro area names, in-migration counts, out-migration counts, and associated state abbreviations, making it suitable for further analysis.

### Preview of Data

```{r}
metro_migration <- metro_to_metro_migration()
head(metro_migration)
dim(metro_migration)
```

## Task 6: Exploratory Data Analysis

### Question 1: Which states have had the highest net population growth rates over the past decade?

::: {.callout-note collapse="true" title="Show Code"}
```{r}
state_growth <- state_population_tidy %>%
  filter(year %in% c(2015, 2024)) %>%
  select(state, year, total_population) %>%
  pivot_wider(names_from = year, values_from = total_population, names_prefix = "pop_") %>%
  mutate(
    growth = pop_2024 - pop_2015,
    growth_rate = growth / pop_2015
  ) %>%
  arrange(desc(growth_rate))

top_growth_state <- state_growth %>% slice(1) %>% pull(state)
top_growth_rate <- state_growth %>% slice(1) %>% pull(growth_rate)

state_growth %>%
  slice_head(n = 10) %>%
  mutate(growth_rate = percent(growth_rate, accuracy = 0.1)) %>%
  gt()
```
:::

The state with the highest population growth rate from 2015 to 2024 was **`r top_growth_state`**, with a growth rate of **`r percent(top_growth_rate, accuracy = 0.1)`**.

### Question 2: Migration Into and Out of New York State

::: {.callout-note collapse="true" title="Show Code"}
```{r}
ny_inflows <- state_flows %>%
  filter(year_current == 2024, state_current == "NY", state_1y != "NY") %>%
  arrange(desc(population))

ny_outflows <- state_flows %>%
  filter(year_current == 2024, state_1y == "NY", state_current != "NY") %>%
  arrange(desc(population))

ny_top_origin <- ny_inflows %>% slice(1) %>% pull(state_1y)
ny_top_dest <- ny_outflows %>% slice(1) %>% pull(state_current)

ny_inflows %>% slice_head(n = 10) %>% gt()
ny_outflows %>% slice_head(n = 10) %>% gt()
```
:::

The most common origin state for individuals moving into New York is **`r ny_top_origin`**, while the most common destination for those leaving New York is **`r ny_top_dest`**.

### Question 3: Migration Into and Out of New York City Metro

::: {.callout-note collapse="true" title="Show Code"}
```{r}
nyc_metro <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"

nyc_in <- metro_migration %>%
  filter(metro_current == nyc_metro, metro_1y != nyc_metro) %>%
  arrange(desc(moved_in))

nyc_out <- metro_migration %>%
  filter(metro_1y == nyc_metro, metro_current != nyc_metro) %>%
  arrange(desc(moved_out))

nyc_top_origin_metro <- nyc_in %>% slice(1) %>% pull(metro_1y)
nyc_top_dest_metro <- nyc_out %>% slice(1) %>% pull(metro_current)

nyc_in %>% select(metro_1y, moved_in) %>% slice_head(n = 10) %>% gt()
nyc_out %>% select(metro_current, moved_out) %>% slice_head(n = 10) %>% gt()
```
:::

The most common metro area people move from into the New York City metro area is **`r nyc_top_origin_metro`**, while the most common destination metro for those leaving the New York City metro area is **`r nyc_top_dest_metro`**.

### Question 4: States with Highest Migration Levels

::: {.callout-note collapse="true" title="Show Code"}
```{r}
state_in_mig <- state_flows %>%
  filter(year_current == 2024, state_current != state_1y) %>%
  group_by(state_current) %>%
  summarise(in_migration = sum(population), .groups = "drop") %>%
  rename(state = state_current)

state_out_mig <- state_flows %>%
  filter(year_current == 2024, state_current != state_1y) %>%
  group_by(state_1y) %>%
  summarise(out_migration = sum(population), .groups = "drop") %>%
  rename(state = state_1y)

state_mig_summary <- state_in_mig %>%
  full_join(state_out_mig, by = "state") %>%
  mutate(
    in_migration = replace_na(in_migration, 0),
    out_migration = replace_na(out_migration, 0),
    net_migration = in_migration - out_migration
  )

state_mig_summary %>% arrange(desc(in_migration)) %>% slice_head(n = 10) %>% gt()
state_mig_summary %>% arrange(desc(out_migration)) %>% slice_head(n = 10) %>% gt()
state_mig_summary %>% arrange(desc(net_migration)) %>% slice_head(n = 10) %>% gt()
```
:::

The summary tables above show the states with the largest total in-migration, total out-migration, and net migration in 2024.

### Question 5: Metro Areas with Highest Migration Levels

::: {.callout-note collapse="true" title="Show Code"}
```{r}
metro_in_summary <- metro_migration %>%
  group_by(metro_current) %>%
  summarise(in_migration = sum(moved_in), .groups = "drop")

metro_out_summary <- metro_migration %>%
  group_by(metro_1y) %>%
  summarise(out_migration = sum(moved_out), .groups = "drop") %>%
  rename(metro_current = metro_1y)

metro_summary <- metro_in_summary %>%
  full_join(metro_out_summary, by = "metro_current") %>%
  mutate(
    in_migration = replace_na(in_migration, 0),
    out_migration = replace_na(out_migration, 0),
    net_migration = in_migration - out_migration
  )

metro_summary %>% arrange(desc(in_migration)) %>% slice_head(n = 10) %>% gt()
metro_summary %>% arrange(desc(out_migration)) %>% slice_head(n = 10) %>% gt()
metro_summary %>% arrange(desc(net_migration)) %>% slice_head(n = 10) %>% gt()
```
:::

The summary tables above show the metro areas with the largest in-migration, out-migration, and net migration totals.

### Question 6: States with Highest Stability

::: {.callout-note collapse="true" title="Show Code"}
```{r}
state_pop_latest <- state_population_tidy %>%
  filter(year == 2024)

stayed_put <- state_flows %>%
  filter(year_current == 2024, state_current == state_1y) %>%
  group_by(state_current) %>%
  summarise(stayed_population = sum(population), .groups = "drop") %>%
  rename(state = state_current) %>%
  left_join(state_pop_latest, by = "state") %>%
  mutate(prop_stayed = stayed_population / total_population) %>%
  arrange(desc(prop_stayed))

stayed_put %>% slice_head(n = 10) %>% gt()
```
:::

The table above shows the states with the highest share of residents who remained in the same state.

### Question 7: Growth Driven by Migration

::: {.callout-note collapse="true" title="Show Code"}
```{r}
pop_growth <- state_population_tidy %>%
  filter(year %in% c(2023, 2024)) %>%
  pivot_wider(names_from = year, values_from = total_population) %>%
  mutate(total_growth = `2024` - `2023`)

growth_vs_migration <- pop_growth %>%
  left_join(state_mig_summary, by = "state") %>%
  mutate(
    frac = if_else(total_growth > 0, net_migration / total_growth, NA_real_)
  ) %>%
  filter(total_growth > 0) %>%
  arrange(desc(frac))

top_migration_growth_state <- growth_vs_migration %>% slice(1) %>% pull(state)

growth_vs_migration %>% slice_head(n = 10) %>% gt()
```
:::

The state with the highest proportion of growth driven by migration is **`r top_migration_growth_state`**.

### Question 8–11: State + Metro Deep Dive (New York)

Largest metro: **New York-Newark-Jersey City, NY-NJ-PA Metro Area**

::: {.callout-note collapse="true" title="Show Code"}
```{r}
focus_state <- "NY"
focus_metro <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"

focus_inflows <- state_flows %>%
  filter(year_current == 2024, state_current == focus_state, state_1y != focus_state)

focus_outflows <- state_flows %>%
  filter(year_current == 2024, state_1y == focus_state, state_current != focus_state)

focus_top_origin_state <- focus_inflows %>% arrange(desc(population)) %>% slice(1) %>% pull(state_1y)
focus_top_dest_state <- focus_outflows %>% arrange(desc(population)) %>% slice(1) %>% pull(state_current)

metro_focus_in <- metro_migration %>%
  filter(metro_current == focus_metro, metro_1y != focus_metro)

metro_focus_out <- metro_migration %>%
  filter(metro_1y == focus_metro, metro_current != focus_metro)

focus_top_origin_metro <- metro_focus_in %>% arrange(desc(moved_in)) %>% slice(1) %>% pull(metro_1y)
focus_top_dest_metro <- metro_focus_out %>% arrange(desc(moved_out)) %>% slice(1) %>% pull(metro_current)

metro_connections <- metro_migration %>%
  filter(metro_1y == focus_metro, metro_current != focus_metro) %>%
  group_by(metro_current) %>%
  summarise(connection = sum(moved_out), .groups = "drop") %>%
  arrange(desc(connection))

metro_connections %>% slice_head(n = 10) %>% gt()
```
:::

For New York, the most common origin state among movers into the state is **`r focus_top_origin_state`**, and the most common destination state among movers leaving New York is **`r focus_top_dest_state`**. At the metro level, the top origin metro feeding into the New York City metro area is **`r focus_top_origin_metro`**, while the top destination metro for movers leaving the region is **`r focus_top_dest_metro`**.

## Task 7: Population Projections

::: {.callout-note collapse="true" title="Show Code"}
```{r}
pop_projection_2030 <- state_population_tidy %>%
  filter(year == 2024) %>%
  mutate(population = total_population * 1.03)

pop_projection_2030 %>%
  arrange(desc(population)) %>%
  select(state, population) %>%
  slice_head(n = 10) %>%
  gt() %>%
  fmt_number(columns = population, decimals = 0)
```
:::

To produce a simple 2030 projection, I applied a modest overall growth factor to the 2024 population values. This provides a basic projected population table for later reapportionment.

## Task 8: Reapportionment

::: {.callout-note collapse="true" title="Show Code"}
```{r}
ny_districts_2030 <- round(
  pop_projection_2030$population[
    pop_projection_2030$state == "New York"
  ] / 750000
)
```
:::

Based on the projected 2030 population, New York is estimated to have **`r ny_districts_2030`** congressional districts.

## Task 9: Developing an Advertising Strategy to Induce Internal Migration and Increase State Political Power

To increase New York’s political power, the strategy should focus on increasing net in-migration while also reducing out-migration. Based on the earlier analysis, New York has strong migration connections with states such as New Jersey, Florida, Texas, and California, as well as major metro areas connected to the New York City region.

To increase in-migration, I would target nearby and connected metro areas where migration to New York is already common. Campaigns would emphasize job opportunities, career growth, transportation access, culture, and long-term economic mobility. To reduce out-migration, I would also target places where former New Yorkers often relocate, using return-focused messaging that highlights New York’s professional and social advantages.

The campaign would rely mainly on digital advertising, social media, influencer partnerships, and transit-based placements in the target metro areas. The most effective audience would likely be young professionals, recent graduates, and remote workers who are more mobile and more responsive to lifestyle and career messaging.

A fitting slogan for this campaign would be **“New York Moves You Forward.”** This message reinforces the idea that New York offers upward mobility, opportunity, and a strong future for residents.

::: {.callout-note collapse="true" title="AI Usage Statement"}
I used generative AI tools (ChatGPT) to assist with debugging R code and structuring Quarto formatting. The AI helped identify syntax errors, improve code organization, and ensure that outputs aligned with the project requirements. All analysis decisions, interpretations, and final written responses were completed and verified by me. I reviewed and modified any AI-generated suggestions to ensure accuracy and alignment with course concepts.
:::