---
title: "STA 9750 Mini Project 4"
author: "Izabelle Diaz"
---

## Task 1: Acquire Olympedia Data

For this task, I collected Olympic medal data directly from Olympedia. To keep the scraping process manageable, I used selected Olympedia edition and sport pages to build the required tidy medal table structure.

::: {.callout-note collapse="true" title="Show Code"}
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
library(stringr)
library(fs)

slow_download <- purrr::slowly(
  download.file,
  purrr::rate_delay(pause = 0.25)
)

read_olympedia <- function(url, ...) {
  
  url <- stringr::str_remove(
    url,
    "https://www.olympedia.org/"
  )
  
  if (stringr::str_starts(url, "/")) {
    url <- stringr::str_remove(url, "/")
  }
  
  cache_dir <- fs::path("data", "mp04", "olympedia_cache")
  
  if (!fs::dir_exists(cache_dir)) {
    fs::dir_create(cache_dir)
  }
  
  cache_path <- fs::path(
    cache_dir,
    url |>
      as.character() |>
      stringr::str_replace_all("/", "_")
  )
  
  if (!fs::file_exists(cache_path)) {
    src_url <- paste0("https://www.olympedia.org/", url)
    
    tryCatch(
      slow_download(src_url, cache_path, mode = "wb"),
      error = function(e) {
        message("Failed to download: ", src_url)
      }
    )
  }
  
  if (!fs::file_exists(cache_path)) {
    return(NULL)
  }
  
  rvest::read_html(cache_path)
}
```
:::

### Selected Olympedia Pages

::: {.callout-note collapse="true" title="Show Code"}
```{r}
sport_links <- tibble(
  games = c(
    "Summer Games of the I Olympiad (1896 Athina)",
    "Summer Games of the II Olympiad (1900 Paris)",
    "Summer Games of the II Olympiad (1900 Paris)",
    "Summer Games of the III Olympiad (1904 St. Louis)",
    "Summer Games of the III Olympiad (1904 St. Louis)"
  ),
  year = c(1896, 1900, 1900, 1904, 1904),
  season = "Summer",
  discipline = c(
    "Shooting",
    "Golf",
    "Swimming",
    "Golf",
    "Artistic Gymnastics"
  ),
  sport_url = c(
    "https://www.olympedia.org/editions/1/sports/SHO",
    "https://www.olympedia.org/editions/2/sports/GLF",
    "https://www.olympedia.org/editions/2/sports/SWM",
    "https://www.olympedia.org/editions/3/sports/GLF",
    "https://www.olympedia.org/editions/3/sports/GAR"
  )
)

glimpse(sport_links)
```
:::

### Scrape Medal Tables

::: {.callout-note collapse="true" title="Show Code"}
```{r}
get_medal_table <- function(sport_url) {
  
  page <- read_olympedia(sport_url)
  
  if (is.null(page)) {
    return(tibble())
  }
  
  tables <- rvest::html_table(
    page,
    fill = TRUE
  )
  
  medal_table <- tables |>
    purrr::keep(
      function(tbl) {
        all(
          c(
            "Gold",
            "Silver",
            "Bronze",
            "Total"
          ) %in% names(tbl)
        )
      }
    )
  
  if (length(medal_table) == 0) {
    return(tibble())
  }
  
  medal_table[[1]] |>
    tibble::as_tibble(.name_repair = "unique") |>
    dplyr::rename(
      competitor = NOC...1
    ) |>
    dplyr::mutate(
      gold = as.integer(Gold),
      silver = as.integer(Silver),
      bronze = as.integer(Bronze),
      total = as.integer(Total)
    ) |>
    dplyr::select(
      competitor,
      gold,
      silver,
      bronze,
      total
    )
}
```
:::

### Create Final Tidy Table

::: {.callout-note collapse="true" title="Show Code"}
```{r message=FALSE, warning=FALSE}
olympic_medals_by_discipline <- sport_links |>
  dplyr::mutate(
    medal_results = purrr::map(
      sport_url,
      get_medal_table
    )
  ) |>
  tidyr::unnest(medal_results) |>
  dplyr::select(
    games,
    year,
    season,
    discipline,
    competitor,
    gold,
    silver,
    bronze,
    total
  )

glimpse(olympic_medals_by_discipline)
```
:::

### Final Tidy Table Check

::: {.callout-note collapse="true" title="Show Code"}
```{r}
set.seed(123)

olympic_medals_by_discipline |>
  dplyr::slice_sample(n = 5)
```
:::

By the end of this step, the selected Olympedia data was combined into a tidy table where each row represents one country’s medal results within a specific discipline and Olympic Games.

## Task 2: Standardize and Clean Countries

For this task, I standardized country names from the Olympedia medal table into ISO-style three-letter country codes. This makes the medal data easier to join with country-level spatial data for mapping.

::: {.callout-note collapse="true" title="Show Code"}
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(countrycode)
library(rnaturalearth)

world_countries <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf"
) |>
  dplyr::select(
    country_name = name,
    iso3 = adm0_a3,
    geometry
  ) |>
  dplyr::filter(
    !is.na(iso3),
    iso3 != "-99"
  )

olympic_country_medals <- olympic_medals_by_discipline |>
  dplyr::mutate(
    iso3 = countrycode(
      competitor,
      origin = "country.name",
      destination = "iso3c"
    )
  ) |>
  dplyr::mutate(
    iso3 = dplyr::recode(
      iso3,
      "URS" = "RUS",
      "GRE" = "GRC"
    )
  )

missing_country_codes <- olympic_country_medals |>
  dplyr::distinct(competitor, iso3) |>
  dplyr::anti_join(
    world_countries,
    by = "iso3"
  )

missing_country_codes
```
:::

### Join Medal Data to Spatial Data

::: {.callout-note collapse="true" title="Show Code"}
```{r}
country_medal_totals <- olympic_country_medals |>
  dplyr::group_by(competitor, iso3) |>
  dplyr::summarise(
    gold = sum(gold, na.rm = TRUE),
    silver = sum(silver, na.rm = TRUE),
    bronze = sum(bronze, na.rm = TRUE),
    total = sum(total, na.rm = TRUE),
    .groups = "drop"
  )

medal_map_data <- world_countries |>
  dplyr::left_join(
    country_medal_totals,
    by = "iso3"
  ) |>
  dplyr::mutate(
    gold = replace_na(gold, 0),
    silver = replace_na(silver, 0),
    bronze = replace_na(bronze, 0),
    total = replace_na(total, 0)
  )

glimpse(medal_map_data)
```
:::

### Geospatial Visualization

::: {.callout-note collapse="true" title="Show Code"}
```{r}
ggplot(medal_map_data) +
  geom_sf(aes(fill = total)) +
  labs(
    title = "Olympic Medal Totals by Country",
    subtitle = "Selected Olympedia sport pages from early Summer Olympic Games",
    fill = "Total Medals",
    caption = "Data scraped from Olympedia and joined to Natural Earth country geometries."
  ) +
  theme_minimal()
```
:::

This map shows how medals from the selected Olympedia pages are distributed across countries. Even in this smaller sample, the United States stands out because of its strong results in early Olympic events such as shooting, golf, and gymnastics.

## Task 3: Exploratory Data Analysis

For this task, I explored patterns in Olympic medal performance using the cleaned Olympedia data. The analysis focused on three main effects: baseline United States performance, the host country effect, and the effect of introducing new sports. Additional exploratory findings were also included to better understand medal distributions across countries and disciplines.

### United States Baseline Performance

The United States appeared frequently in the selected Olympedia sample and earned some of the highest medal totals overall.

#### Finding 1: United States Medal Totals

::: {.callout-note collapse="true" title="Show Code"}
```{r}
usa_summary <- olympic_medals_by_discipline |>
  dplyr::filter(competitor == "United States") |>
  dplyr::summarise(
    total_gold = sum(gold, na.rm = TRUE),
    total_silver = sum(silver, na.rm = TRUE),
    total_bronze = sum(bronze, na.rm = TRUE),
    total_medals = sum(total, na.rm = TRUE)
  )

usa_summary
```
:::

The United States earned a total of `r usa_summary$total_medals` medals in the selected Olympic sample.

#### Finding 2: United States Medal Distribution by Sport

::: {.callout-note collapse="true" title="Show Code"}
```{r}
usa_sports <- olympic_medals_by_discipline |>
  dplyr::filter(competitor == "United States") |>
  dplyr::group_by(discipline) |>
  dplyr::summarise(
    total_medals = sum(total, na.rm = TRUE),
    .groups = "drop"
  )

usa_sports
```
:::

#### Finding 3: United States Visualization

::: {.callout-note collapse="true" title="Show Code"}
```{r}
ggplot(usa_sports,
       aes(
         x = reorder(discipline, total_medals),
         y = total_medals
       )) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "United States Medal Totals by Discipline",
    x = "Discipline",
    y = "Total Medals"
  ) +
  theme_minimal()
```
:::

The United States performed especially well in Artistic Gymnastics, followed by golf and shooting, in this early Olympic sample.

---

### Host Country Effect

Hosting the Olympic Games may provide advantages such as familiarity with the environment, larger team representation, and home crowd support.

#### Finding 4: Host Countries in the Sample

::: {.callout-note collapse="true" title="Show Code"}
```{r}
host_countries <- tibble(
  year = c(1896, 1900, 1904),
  host_country = c("Greece", "France", "United States")
)

host_countries
```
:::

#### Finding 5: Host Country Medal Totals

::: {.callout-note collapse="true" title="Show Code"}
```{r}
host_effect <- olympic_medals_by_discipline |>
  dplyr::left_join(
    host_countries,
    by = "year"
  ) |>
  dplyr::mutate(
    is_host = competitor == host_country
  ) |>
  dplyr::group_by(is_host) |>
  dplyr::summarise(
    avg_medals = mean(total, na.rm = TRUE),
    total_medals = sum(total, na.rm = TRUE),
    .groups = "drop"
  )

host_effect
```
:::

Host countries tended to earn more medals on average than non-host countries in the selected sample.

#### Finding 6: Host Country Visualization

::: {.callout-note collapse="true" title="Show Code"}
```{r}
ggplot(host_effect,
       aes(
         x = is_host,
         y = avg_medals,
         fill = is_host
       )) +
  geom_col() +
  labs(
    title = "Average Medal Totals: Host vs Non-Host Countries",
    x = "Host Country",
    y = "Average Medals"
  ) +
  theme_minimal()
```
:::

#### Finding 7: Host Country Statistical Test

::: {.callout-note collapse="true" title="Show Code"}
```{r}
library(infer)

host_test_data <- olympic_medals_by_discipline |>
  dplyr::left_join(
    host_countries,
    by = "year"
  ) |>
  dplyr::mutate(
    is_host = if_else(
      competitor == host_country,
      "Host",
      "Non-Host"
    )
  )

host_t_test <- host_test_data |>
  infer::t_test(
    total ~ is_host,
    order = c("Non-Host", "Host")
  )

host_t_test
```
:::

This test evaluates whether host countries earned significantly different medal totals compared to non-host countries.

---

### New Sport Effect

Different sports contributed unevenly to medal totals. Some sports had especially high medal concentrations among a few countries.

#### Finding 8: Medal Totals by Discipline

::: {.callout-note collapse="true" title="Show Code"}
```{r}
discipline_summary <- olympic_medals_by_discipline |>
  dplyr::group_by(discipline) |>
  dplyr::summarise(
    total_medals = sum(total, na.rm = TRUE),
    countries = n_distinct(competitor),
    .groups = "drop"
  )

discipline_summary
```
:::

#### Finding 9: Discipline Visualization

::: {.callout-note collapse="true" title="Show Code"}
```{r}
ggplot(
  discipline_summary,
  aes(
    x = reorder(discipline, total_medals),
    y = total_medals
  )
) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Total Olympic Medals by Discipline",
    x = "Discipline",
    y = "Total Medals"
  ) +
  theme_minimal()
```
:::

Artistic Gymnastics produced the largest medal totals among the selected disciplines.

#### Finding 10: Countries Participating by Discipline

::: {.callout-note collapse="true" title="Show Code"}
```{r}
discipline_summary |>
  dplyr::select(
    discipline,
    countries
  )
```
:::

Some sports involved fewer participating countries, which may increase medal concentration among dominant nations.

#### Finding 11: Proportion Test

::: {.callout-note collapse="true" title="Show Code"}
```{r}
sport_prop_data <- olympic_medals_by_discipline |>
  dplyr::mutate(
    high_medal_total = total >= 3,
    gymnastics = discipline == "Artistic Gymnastics"
  )

gymnastics_attempts <- sport_prop_data |>
  dplyr::filter(gymnastics) |>
  nrow()

gymnastics_wins <- sport_prop_data |>
  dplyr::filter(gymnastics) |>
  dplyr::summarise(
    wins = sum(high_medal_total, na.rm = TRUE)
  ) |>
  dplyr::pull(wins)

sport_prop_test <- broom::tidy(
  stats::binom.test(
    x = gymnastics_wins,
    n = gymnastics_attempts
  )
)

sport_prop_test
```
:::

This exact binomial test evaluates whether Artistic Gymnastics medal rows were likely to contain high medal totals in the selected sample.


### Additional Exploratory Findings

#### Finding 12: Top Countries by Total Medals

::: {.callout-note collapse="true" title="Show Code"}
```{r}
top_countries <- olympic_medals_by_discipline |>
  dplyr::group_by(competitor) |>
  dplyr::summarise(
    total_medals = sum(total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(desc(total_medals))

top_countries
```
:::

Germany and the United States were among the strongest performers in the selected Olympic sample. 

## Task 4: Statistical Inference

::: {.callout-note collapse="true" title="Show Code"}
```{r}
library(tidyverse)
library(infer)
library(broom)

medal_types <- c(
  "gold",
  "silver",
  "bronze",
  "total"
)

host_countries <- tibble(
  year = c(1896, 1900, 1904),
  season = "Summer",
  host_country = c(
    "Greece",
    "France",
    "United States"
  )
)

usa_yearly_medals <- olympic_medals_by_discipline |>
  filter(
    competitor == "United States",
    season == "Summer"
  ) |>
  group_by(year) |>
  summarise(
    gold = sum(gold, na.rm = TRUE),
    silver = sum(silver, na.rm = TRUE),
    bronze = sum(bronze, na.rm = TRUE),
    total = sum(total, na.rm = TRUE),
    .groups = "drop"
  )

usa_baseline_ci <- purrr::map_dfr(
  medal_types,
  function(medal) {
    
    usa_yearly_medals |>
      infer::t_test(
        as.formula(
          paste(medal, "~ NULL")
        )
      ) |>
      mutate(
        medal_type = medal,
        quantity = "USA baseline average"
      )
  }
)

usa_baseline_ci
```
:::

### Host Country Effect

::: {.callout-note collapse="true" title="Show Code"}
```{r}
country_yearly_medals <- olympic_medals_by_discipline |>
  group_by(
    year,
    season,
    competitor
  ) |>
  summarise(
    gold = sum(gold, na.rm = TRUE),
    silver = sum(silver, na.rm = TRUE),
    bronze = sum(bronze, na.rm = TRUE),
    total = sum(total, na.rm = TRUE),
    .groups = "drop"
  )

host_medal_data <- host_countries |>
  left_join(
    country_yearly_medals,
    by = c(
      "year",
      "season",
      "host_country" = "competitor"
    )
  ) |>
  arrange(season, year)

host_prior_data <- host_medal_data |>
  select(
    year,
    season,
    host_country,
    all_of(medal_types)
  ) |>
  pivot_longer(
    cols = all_of(medal_types),
    names_to = "medal_type",
    values_to = "host_current"
  ) |>
  left_join(
    country_yearly_medals |>
      select(
        prior_year = year,
        season,
        competitor,
        all_of(medal_types)
      ) |>
      pivot_longer(
        cols = all_of(medal_types),
        names_to = "medal_type",
        values_to = "host_prior"
      ),
    by = c(
      "season",
      "host_country" = "competitor",
      "medal_type"
    )
  ) |>
  filter(prior_year < year) |>
  group_by(
    year,
    season,
    host_country,
    medal_type
  ) |>
  slice_max(
    prior_year,
    n = 1,
    with_ties = FALSE
  ) |>
  ungroup() |>
  filter(
    !is.na(host_prior),
    host_prior > 0
  ) |>
  mutate(
    percent_increase =
      (host_current - host_prior) /
      host_prior,
    
    host_factor =
      host_current / host_prior
  )

host_effect_ci <- host_prior_data |>
  group_by(medal_type) |>
  summarise(
    estimate =
      mean(
        percent_increase,
        na.rm = TRUE
      ),
    
    lower_ci =
      estimate -
      1.96 *
      sd(
        percent_increase,
        na.rm = TRUE
      ) /
      sqrt(n()),
    
    upper_ci =
      estimate +
      1.96 *
      sd(
        percent_increase,
        na.rm = TRUE
      ) /
      sqrt(n()),
    
    n = n(),
    quantity =
      "Host country percent increase",
    
    .groups = "drop"
  )

host_effect_ci
```
:::

Because the reduced sample only produced one valid host-country comparison for some medal types, confidence intervals could not be calculated and therefore appear as NA.

### New Sport Effect

::: {.callout-note collapse="true" title="Show Code"}
```{r}
discipline_years <- olympic_medals_by_discipline |>
  distinct(
    year,
    season,
    discipline
  ) |>
  arrange(
    season,
    discipline,
    year
  ) |>
  group_by(
    season,
    discipline
  ) |>
  mutate(
    previous_year = lag(year),
    is_new_sport = is.na(previous_year)
  ) |>
  ungroup()

new_sport_host_data <- olympic_medals_by_discipline |>
  left_join(
    discipline_years,
    by = c(
      "year",
      "season",
      "discipline"
    )
  ) |>
  left_join(
    host_countries,
    by = c(
      "year",
      "season"
    )
  ) |>
  filter(
    is_new_sport,
    competitor == host_country
  )

new_sport_probability_ci <- purrr::map_dfr(
  medal_types,
  function(medal) {
    
    test_data <- new_sport_host_data |>
      dplyr::mutate(
        host_won =
          .data[[medal]] > 0
      )
    
    wins <- sum(
      test_data$host_won,
      na.rm = TRUE
    )
    
    attempts <- nrow(test_data)
    
    if (attempts == 0) {
      
      return(
        tibble(
          statistic = NA_real_,
          chisq_df = NA_real_,
          p_value = NA_real_,
          alternative = "two.sided",
          lower_ci = NA_real_,
          upper_ci = NA_real_,
          medal_type = medal,
          quantity =
            "Probability host wins medal in new sport"
        )
      )
    }
    
    broom::tidy(
      stats::binom.test(
        x = wins,
        n = attempts
      )
    ) |>
      dplyr::transmute(
        statistic = estimate,
        chisq_df = NA_real_,
        p_value = p.value,
        alternative = alternative,
        lower_ci = conf.low,
        upper_ci = conf.high,
        medal_type = medal,
        quantity =
          "Probability host wins medal in new sport"
      )
  }
)

new_sport_probability_ci
```
:::

Overall, these confidence intervals estimate the United States baseline medal performance, the expected host-country boost, and the probability that host countries win medals in newly introduced Olympic sports. Each analysis was repeated for gold, silver, bronze, and total medals.

## Task 5: Combining Confidence Intervals to Forecast US Medal Counts

For the final forecast, I combined the confidence intervals from Task 4 using a Monte Carlo simulation. This follows the assignment’s simplified assumption that the intervals are independent.

::: {.callout-note collapse="true" title="Show Code"}
```{r}
set.seed(123)

host_effect_complete <- host_effect_ci |>
  mutate(
    lower_ci = if_else(
      is.na(lower_ci),
      estimate,
      lower_ci
    ),
    upper_ci = if_else(
      is.na(upper_ci),
      estimate,
      upper_ci
    )
  )

forecast_inputs <- usa_baseline_ci |>
  select(
    medal_type,
    baseline_estimate = estimate,
    baseline_lower = lower_ci,
    baseline_upper = upper_ci
  ) |>
  left_join(
    host_effect_complete |>
      mutate(
        host_factor_estimate = 1 + estimate,
        host_factor_lower = 1 + lower_ci,
        host_factor_upper = 1 + upper_ci
      ) |>
      select(
        medal_type,
        host_factor_estimate,
        host_factor_lower,
        host_factor_upper
      ),
    by = "medal_type"
  ) |>
  left_join(
    new_sport_probability_ci |>
      select(
        medal_type,
        new_sport_lower = lower_ci,
        new_sport_upper = upper_ci
      ),
    by = "medal_type"
  ) |>
  mutate(
    host_factor_lower = replace_na(
      host_factor_lower,
      1
    ),
    host_factor_upper = replace_na(
      host_factor_upper,
      1
    ),
    new_sport_lower = replace_na(
      new_sport_lower,
      0
    ),
    new_sport_upper = replace_na(
      new_sport_upper,
      0
    )
  )

forecast_inputs
```
:::

::: {.callout-note collapse="true" title="Show Code"}
```{r}
simulate_forecast <- function(
    baseline_lower,
    baseline_upper,
    host_factor_lower,
    host_factor_upper,
    new_sport_lower,
    new_sport_upper,
    sims = 10000
) {
  
  baseline_mean <-
    (baseline_lower + baseline_upper) / 2
  
  host_factor_mean <-
    (host_factor_lower + host_factor_upper) / 2
  
  new_sport_mean <-
    (new_sport_lower + new_sport_upper) / 2
  
  baseline_sd <- abs(
    baseline_upper - baseline_mean
  ) / qnorm(0.975)
  
  host_factor_sd <- abs(
    host_factor_upper - host_factor_mean
  ) / qnorm(0.975)
  
  new_sport_sd <- abs(
    new_sport_upper - new_sport_mean
  ) / qnorm(0.975)
  
  baseline_samples <- rnorm(
    sims,
    mean = baseline_mean,
    sd = baseline_sd
  )
  
  host_factor_samples <- rnorm(
    sims,
    mean = host_factor_mean,
    sd = host_factor_sd
  )
  
  new_sport_samples <- rnorm(
    sims,
    mean = new_sport_mean,
    sd = new_sport_sd
  )
  
  forecast_samples <-
    baseline_samples *
    host_factor_samples +
    22 * new_sport_samples
  
  forecast_samples <- pmax(
    forecast_samples,
    0
  )
  
  tibble(
    forecast_estimate =
      mean(
        forecast_samples,
        na.rm = TRUE
      ),
    
    lower_ci =
      quantile(
        forecast_samples,
        0.025,
        na.rm = TRUE
      ),
    
    upper_ci =
      quantile(
        forecast_samples,
        0.975,
        na.rm = TRUE
      )
  )
}
```
:::

::: {.callout-note collapse="true" title="Show Code"}
```{r}
usa_2028_forecast <- forecast_inputs |>
  mutate(
    forecast = pmap(
      list(
        baseline_lower,
        baseline_upper,
        host_factor_lower,
        host_factor_upper,
        new_sport_lower,
        new_sport_upper
      ),
      simulate_forecast
    )
  ) |>
  unnest(forecast)

usa_2028_forecast |>
  dplyr::select(
    medal_type,
    forecast_estimate,
    lower_ci,
    upper_ci
  )
```
:::

The forecast combines three pieces of evidence: Team USA’s baseline medal performance, the estimated host nation factor, and the expected contribution from 22 additional medal events in 2028. Because this report uses a reduced Olympedia sample for rendering purposes, these estimates should be interpreted as a simplified demonstration of the Monte Carlo forecasting method, not as literal medal-count predictions.

Task 6: Team USA Fundraising Fact Sheet

## Team USA LA 2028 Olympic Forecast Report

### Overview

As the United States prepares to host the 2028 Summer Olympic Games in Los Angeles, expectations for Team USA performance remain high. Using Olympedia medal data, exploratory analysis, statistical inference, and Monte Carlo simulation methods, this report forecasts Team USA’s medal outlook for LA 2028 and highlights the major drivers behind projected success.

The forecasting model used in this report combines three major factors:

- Team USA’s historical baseline Olympic performance
- The historical host-country performance effect
- The impact of newly added Olympic events

Together, these factors suggest that Team USA is positioned to remain one of the strongest Olympic programs entering LA 2028.

### Factor 1: Team USA Baseline Performance

Historical Olympic results show that the United States consistently performs well across multiple Summer Olympic disciplines. In the selected Olympedia sample, Team USA earned strong medal totals in sports such as Artistic Gymnastics, golf, and shooting, reinforcing the country’s long-standing depth across Olympic competition.

Confidence intervals constructed using the infer package estimated Team USA’s average medal performance across gold, silver, bronze, and total medals. While the reduced sample size creates wider intervals, the overall trend still indicates strong baseline competitiveness.

#### Baseline Medal Estimates

| Medal Type | Average Estimate |
|---|---|
| Gold | 6.0 |
| Silver | 4.0 |
| Bronze | 4.3 |
| Total | 14.3 |

These baseline estimates serve as the foundation of the 2028 medal forecast.

### Factor 2: Host Country Advantage

Host nations historically experience performance increases during Olympic Games hosted on home soil. Advantages may include larger athlete delegations, familiar environments, reduced travel strain, and increased crowd support.

The analysis compared host-country medal totals against their performance in the prior same-season Olympics. Even within the reduced sample, host nations generally experienced positive medal increases.

For Team USA, hosting the 2028 Olympics in Los Angeles may provide a meaningful competitive advantage relative to non-host Olympic years.

#### Estimated Host Effect

| Medal Type | Estimated Percent Increase |
|---|---|
| Gold | Approximately 600% |
| Total | Approximately 1800% |

Because the reduced sample only produced a limited number of valid host-country comparisons, these values should be interpreted as directional indicators rather than precise estimates. However, the overall relationship still supports the historical presence of a host-country advantage.

### Factor 3: New Olympic Events

The LA 2028 Olympics will feature additional medal events compared to previous Games. Historically, newly introduced sports and events create opportunities for host nations to expand medal totals.

Using binomial confidence intervals, the analysis estimated the probability that host countries win medals in newly introduced sports. Results suggested relatively strong probabilities across medal categories, especially for bronze and total medals.

This effect becomes particularly important because LA 2028 is expected to feature 22 additional medal events compared with Paris 2024.

### Monte Carlo Forecast for LA 2028

The final medal forecast combined all three factors using Monte Carlo simulation methods. This approach repeatedly sampled values from the confidence intervals estimated earlier in the report to generate a distribution of possible medal outcomes.

#### Forecasted Team USA Medal Counts

| Medal Type | Forecast Estimate | Lower 95% CI | Upper 95% CI |
|---|---|---|---|
| Gold | 61.2 | 0.0 | 175.0 |
| Silver | 17.1 | 0.0 | 34.4 |
| Bronze | 20.0 | 0.2 | 39.9 |
| Total | 373.0 | 0.0 | 1224.0 |

Because this report uses a reduced Olympedia sample for rendering purposes, the forecast should be interpreted as a demonstration of the modeling approach rather than a literal medal-count prediction.

### Conclusion

Team USA enters the LA 2028 Olympics with several favorable conditions: historically strong baseline performance, the advantages associated with hosting the Games, and opportunities created by newly added Olympic events.

While this analysis uses a simplified forecasting framework, the combined evidence suggests that Team USA is well positioned to remain one of the premier Olympic delegations in the world.

For sponsors and fundraising partners, LA 2028 represents an opportunity to align with a globally recognized athletic program expected to perform at a highly competitive level on home soil. The combination of athletic success, global media visibility, and domestic fan engagement makes Team USA a strong platform for corporate partnership and Olympic investment.

::: {.callout-note title="AI Usage Statement" collapse="true"}

ChatGPT was used to help debug code for this mini-project. All non-code written analysis, interpretations, and final report text were reviewed and edited by the student before submission.

:::